My title

### 2024-09-24 paper/paper.qmd
paper discussion section added

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

@@

@@

@@

Weaknesses and next steps should also be included.

>>>>

The results highlight Toronto's phased approach to developing its cycling infrastructure, with a significant spike in installations in the early 2000s, as shown in @fig-installed. This surge is likely because the dataset started recording from 2001, hence all installations before was counted towards 2001. Futher, the fact that goverment start taking this record and the large number of installations at this period may have reflected political will and public demand for greener transportation alternatives, driven by growing concerns about climate change and urban congestion.

However, after the initial surge, the number of bikeways installed annually began to decline, a pattern that persists to this day. Several factors could explain this decline. First, the initial phase of the bike plan likely focused on areas where cycling infrastructure could be installed with relative ease, particularly in downtown Toronto, where road networks and urban density supported such investments. As the network expanded into more complex areas, such as suburban neighborhoods with fewer existing roadways, the pace of installation slowed. Additionally, the physical limitations of available space in a highly developed urban area may have constrained further expansion, as new bikeways would require more complicated and costly measures, such as road reallocation or shared-use paths.

Despite this slowdown in new installations, @fig-upgraded demonstrates a significant focus on upgrading existing infrastructure, particularly in later years. This shift from installation to upgrades could indicate a strategic pivot by the city toward improving the quality and safety of existing cycling infrastructure. As infrastructure ages and new safety standards emerge, upgrading older lanes becomes crucial to ensuring the network remains usable and safe for cyclists. The peak of upgrades in 2015, for example, might reflect the city's response to increased cycling activity, demands from advocacy groups, or the implementation of updated cycling policies that prioritize safer and more modern designs, such as protected bike lanes.

This pattern of prioritizing upgrades over new installations also suggests that Toronto's initial expansion may have left some areas with suboptimal or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes that offer minimal protection to cyclists. As urban transportation planning evolves, the focus has shifted toward creating more separated and protected lanes, which are essential for encouraging cycling among less experienced riders, families, or those concerned about safety. This ongoing commitment to improving the infrastructure—rather than merely expanding it—reflects a maturation in Toronto’s approach to active transportation, as planners recognize that quality and safety are just as important as quantity when promoting cycling as a viable mode of transport.

Finally, another crucial aspect linked to the decline in installations could be budgetary constraints or competing infrastructure projects. As urban budgets are stretched across various priorities—public transit expansion, road maintenance, and green space development—the pace of installing new cycling lanes might slow. However, the investment in upgrades suggests that cycling infrastructure remains a priority, even if the focus has shifted from expansion to refinement. Continuing to monitor and analyze these trends will be essential for understanding how Toronto can balance its need for both new installations and the improvement of existing cycling routes to meet its long-term sustainability and mobility objectives.

In conclusion, the first discussion point highlights the evolution of Toronto’s cycling infrastructure from an initial phase of rapid expansion to a more recent focus on upgrading and enhancing the safety and quality of the network. This shift reflects the challenges of balancing growth with maintenance, the need to adapt to changing urban landscapes, and the influence of global best practices and local advocacy in shaping the city’s cycling future.

@@

The comfort level classification of Toronto's cycling infrastructure offers valuable insights into the quality of the city's bike network, particularly with respect to cyclist safety and usability. As shown in @fig-comfort-one, the majority of bikeways fall into the Low or High Comfort categories, with Moderate Comfort lanes—typically those offering relatively protected infrastructure—making up a smaller proportion of the overall network. This distribution suggests that, Toronto has successfully expanded the sheer number of bikeways, the majority could offer the protection and comfort necessary to encourage broader and more consistent usage. However, it is better to see Low Comfort level lanes taking the smallest portion among all.

The classification of cycling lanes into different comfort levels is important because it directly affects who feels confident enough to cycle. Low Comfort lanes, often painted bike lanes or shared roadways, offer minimal physical separation from motor vehicles, making them less appealing to risk-averse cyclists. The lack of protection increases the perceived danger, discouraging people from cycling regularly, particularly in areas with high traffic volumes. For example, a cyclist may feel comfortable using a Low Comfort lane in a quiet residential neighborhood but avoid it entirely on a busy downtown street where traffic and intersections pose greater hazards.

Moderate Comfort lanes, which typically include designated bike lanes, offer more safety but still lack the physical separation provided by protected or High Comfort infrastructure. These lanes are a step in the right direction but may not sufficiently address concerns about safety, particularly in areas with fast-moving or congested vehicle traffic. Moderate Comfort lanes are likely to attract more cyclists than Low Comfort lanes, but they may not be enough to significantly increase ridership across all demographic groups. For instance, studies have shown that women, children, and older adults are particularly sensitive to perceived danger in cycling, and without sufficient protection, these groups are less likely to use the infrastructure regularly.

High Comfort lanes, on the other hand, provide the greatest potential for encouraging cycling across a broad range of demographics. These lanes, often in the form of protected bike lanes or multi-use paths, offer a physically separated space for cyclists, reducing the likelihood of accidents involving motor vehicles. Research from cities that have invested heavily in High Comfort infrastructure, such as Amsterdam or Copenhagen, indicates that the availability of such protected lanes is a key factor in increasing overall cycling rates and making cycling a more viable and appealing transportation option for all. In Toronto, however, as the analysis shows, only a small proportion of the network is classified as High Comfort. This limits the extent to which the city can fully leverage its cycling infrastructure to promote widespread, sustained usage.

The low proportion of High Comfort lanes can be partly attributed to the early phases of Toronto's cycling infrastructure development, during which the focus was likely on quickly installing a large number of lanes to meet political and public demands. At that time, lower-cost solutions like painted bike lanes and sharrows (shared roadways) were easier and faster to implement than more expensive, space-consuming, and structurally complex protected lanes. However, these lanes did not necessarily provide the level of safety needed to support a significant increase in cycling activity, particularly among more cautious cyclists. The subsequent shift towards upgrading infrastructure, as discussed earlier, is likely an acknowledgment of the limitations of the initial approach, where quantity may have been prioritized over quality.

Moreover, the analysis reveals that many of Toronto's Moderate and Low Comfort lanes have yet to be upgraded, as shown in @fig-comfort-one-2 and @fig-comfort-two. For both Low and Moderate Comfort lanes, over 75% have not been upgraded, which implies that much of the city's existing cycling network still operates under outdated or less protective standards. This points to a significant gap between the current state of Toronto's cycling infrastructure and the standards that are increasingly being recognized as necessary to make cycling a mainstream mode of transportation in urban environments. The failure to upgrade a larger portion of these lanes may stem from budgetary constraints, political will, or challenges in reallocating road space in already congested or car-dominated areas.

One of the most critical implications of these findings is that Toronto’s cycling infrastructure might not yet be sufficiently robust to meet the city’s broader goals of reducing car dependency, lowering emissions, and promoting healthier, more active lifestyles. To truly shift a significant portion of the population towards cycling, the city will need to invest more heavily in upgrading its infrastructure, particularly in transitioning more of its Low and Moderate Comfort lanes into High Comfort lanes. This is especially important in areas where traffic volume and speed are higher, as these conditions pose greater risks to cyclists and deter people from choosing bikes as a mode of transportation.

Another potential reason for the focus on Moderate and Low Comfort lanes is the spatial distribution of the city’s bikeways. As mentioned earlier when showing the result of @fog-map, much of Toronto’s cycling network is concentrated in the downtown core, where road space is at a premium. Installing protected lanes in these areas often requires taking road space away from cars, which can be politically and technically challenging. In contrast, suburban areas, where road space is more abundant, could offer better opportunities for the city to build High Comfort infrastructure. Expanding the reach of the cycling network into these suburban areas, while upgrading the comfort levels of existing downtown lanes, could have a transformative effect on overall cycling rates and help make cycling a viable option for longer commutes.

In conclusion, while Toronto has made considerable progress in expanding its cycling infrastructure, the comfort level analysis reveals that much of the network remains suboptimal for ensuring safety and encouraging more widespread use. To achieve its sustainability and transportation goals, the city must focus on upgrading its existing Low and Moderate Comfort lanes and prioritize the construction of High Comfort infrastructure. This shift will be crucial not only for improving safety but also for making cycling an attractive option for a more diverse range of residents, thereby promoting equity and inclusivity in urban transportation.

@@

The geospatial analysis of Toronto's cycling infrastructure, as visualized in @fig-map, reveals clear spatial patterns that highlight both the strengths and weaknesses of the city’s current bike network. One of the key findings is the dense concentration of bikeways in the downtown core, where cycling infrastructure is most integrated into the existing urban fabric. This concentration aligns with expectations, as downtown Toronto is the city’s most densely populated and heavily trafficked area, where the demand for alternative transportation modes is highest. The high density of bikeways in the city center reflects a concerted effort to provide accessible cycling options in areas where they are most likely to be used, particularly by commuters and urban residents.

However, the map also exposes a noticeable gap in cycling infrastructure as one moves further from the city center. The relative sparsity of bike lanes in suburban and peripheral areas suggests that while downtown residents enjoy a well-connected cycling network, those living in the suburbs have fewer options for safe, comfortable bike routes. This discrepancy may limit the overall effectiveness of Toronto’s cycling infrastructure, as it primarily serves those in the urban core while potentially neglecting a significant portion of the population. Suburban residents, who often have longer commutes and fewer public transit options, could benefit greatly from expanded cycling infrastructure that offers a viable alternative to car travel.

The lack of infrastructure in suburban areas could be attributed to a variety of factors. Suburbs typically have lower population densities and wider roads, making cycling infrastructure less of a priority in urban planning compared to the densely populated downtown core. Additionally, suburban regions tend to have more car-dependent cultures, where cycling is not as ingrained as a common mode of transportation. The existing infrastructure may therefore reflect historical priorities that favored cars over active transportation. However, as cities like Toronto strive to reduce traffic congestion and meet climate goals, expanding cycling infrastructure into these suburban areas could be a crucial step toward encouraging more sustainable transportation choices.

Furthermore, the map shows the importance of major recreational areas, such as parks and waterfronts, in Toronto’s cycling network. Multi-use trails, which provide High Comfort lanes, are particularly prevalent in these areas, offering scenic and safe routes for cyclists. While these trails are beneficial for recreational cycling, they may not be fully integrated into the city’s transportation network, limiting their utility for commuters. Connecting these trails to the downtown core and other major employment hubs could improve their role in promoting cycling as a legitimate form of daily transportation rather than just a leisure activity.

Lastly, the map highlights potential future opportunities for infrastructure expansion. Areas with sparse cycling routes could be prioritized for new installations, especially where there is existing road capacity or public demand for safer cycling options. Identifying and addressing these gaps will be crucial as Toronto continues to grow and seeks to make its transportation system more equitable and sustainable. The current infrastructure is heavily concentrated in downtown areas, but future developments could focus on creating a more geographically balanced network, ensuring that all residents, regardless of location, have access to safe and convenient cycling routes.

In summary, the geospatial distribution of Toronto’s cycling infrastructure reveals a well-connected downtown network but highlights the need for more investment in suburban areas and better integration of recreational trails into the broader transportation system. Expanding the reach of cycling infrastructure into underdeveloped areas will be key to making cycling a viable transportation option for a larger portion of the city’s population.

@@

One key limitation of this study is the reliance on the 'Cycling Network' dataset from OpenDataToronto, which, while comprehensive, might not capture informal or emerging cycling routes that are not officially recognized. Additionally, the dataset may not include newer infrastructure developments if updates were delayed in publication. Future research should integrate real-time data collection techniques or alternative datasets (e.g., crowdsourced information) to ensure that the analysis reflects the most current state of the network.

Another weakness lies in the classification of comfort levels, which, while useful, is a simplification. Comfort in cycling can be influenced by numerous factors beyond infrastructure type, such as traffic density, road conditions, and seasonal weather. A more nuanced classification that incorporates these additional factors could provide a more accurate assessment of how comfortable Toronto's cycling network truly is for different types of cyclists.

Next steps should include focusing on identifying areas with fewer bike lanes and lower comfort levels, particularly in suburban areas. These regions could be targeted for infrastructure upgrades and expansion to promote greater equity in access to safe cycling options. Additionally, analyzing cycling usage patterns, perhaps through mobile GPS data or surveys, would offer valuable insights into how the infrastructure is being utilized, helping city planners to make data-driven decisions in future expansions or upgrades.
